\documentstyle[tengtex]{article}
\pagestyle{empty}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}
\begin{document}
\quenya

I hereby offer these spells to all who would have them, for toil or
for~mirth, and I grant them leave to copy them. But credit they shall
give to me, Tolkien, F\"eanor, Aul\"e and Eru, and they shall not
trade the spells, nor any thing wrought therewith, for gold, silver
and jewels. Death in torment and darkness shall be the doom of the
greedy one who heedeth not this oath, and his name shall not
be~uttered, but shall be forgotten for~ever between the borders of
the world and beyond them.

\begin{elvish}
  sinome antan kuruquettar sine ilyain i merir te mootalen ar
  tyalien, ar lavin ten neunakarie te. nan hantale antavalto
  nin, tolkInen, feanAron, aulen ar iluuvatAren, ar avalto manka
  i kuruquettar ar ilya karna tenen an malta, tyelpe ar mIri.
  qualme i morniesse yeeva umbar i milko i Uva tire sina vanda,
  ar Uva quenta esserya, nA oiale vanwievAro enyaliello imbe
  i rImar ardave ar pella te.
\end{elvish}

\it Sinome antan kuruquettar sine ilyain i merir te m\'otalen ar
tyalien, ar lavin ten neunakarie te. Nan hantale antavalto nin,
Tolk\'\i nen, Fean\'aron, Aulen ar Il\'uvat\'aren, ar avalto manka
i kuruquettar ar ilya karna tenen an malta, tyelpe ar m\'\i ri.
Qualme i morniesse y\'eva umbar i milko i \'uva tire sina vanda,
ar \'uva quenta esserya, n\'a oiale vanwiev\'aro enyaliello imbe
i r\'\i mar Ardave ar pella~te.
\end{document}
